---
title: "Causation:Fisher's Exact P-values"
output: html_notebook
---

In this notebook we shall examine three attempts to pin down causal relationships in social science. We first examine Fisher's approach using exact p-values

### Fisher's Exact p-values and Randomised distributions

The fundamental problem of causal inference is best illustrated within the ideal of a 
randomised experiment where we can observe for each patient their previous traits, 
whether or not they were treated and their individual outcomes. 
It is easy to see that we are unable to observe the counterfactual outcome for each particular 
patient under this treatment plan. A treatment plan $\textbf{ W }$ counts as a randomised experiment 
if each patient is assigned a treatment status with probability between 0 and 1 
according to a function known by the experimenter.  

```{r warning = FALSE, message=FALSE, warning=FALSE, echo= FALSE}
library(MatchIt)
library(xtable)
library(knitr)
library(data.table)
library(DT)

data(lalonde)
medical_example <- data.frame(
  unit <- c("Patient #1", "Patient #2", "Patient #3", "Patient #4", "Patient #5", "Patient  #6"),
  control_outcome <- c(NA, NA, NA, 4, 0, 1),
  treatment_outcome <- c(3, 5, 0, NA, NA, NA),
  treated <- c(1, 1, 1, 0, 0, 0),
  covariates <- c("3, 2, 1", "2, 4, 1", "5, 2, 4", "6, 1, 3", "4, 3, 2", "4, 1, 2")
)
colnames(medical_example) <- c("unit", "Y(0)", "Y(1)", "W", "covariates")
datatable(medical_example)

```

In this table we can see only the outcomes of the experiment for six patients, 
but we are unable to calculate the causal effect at the patient level because we 
will never observe the counterfactual result of not treating (treating) a treated (not treated) patient. 
Instead, following Fisher we seek to observe the average causal effect of the treatment at a population 
level by deriving a test-statistic from our observed results. 

$T(\textbf{W}, \textbf{Y}^{obs}) = T^{diff} = |\overline{Y}^{obs}_{1} - \overline{Y}^{obs}_{0}|
 = |8 / 3 - 5 /3 | = 1.00$

Now this result is insufficient to conclude anything substantial 
but consider the addition of a second assumption regarding the null hypothesis. 
If the null hypothesis is true, (i.e. if the treatment has no effect), then 
we have the following situation where: 

$Y_{c}(0) = Y_{t}(1) = Y^{obs}$ 

holds regardless of treatment plan. Hence we can vary the treatment in the following fashion and calculate 
the test statistics from each. There are ${6 \choose 3} = 20$ treatment plans, 
but we show the first 6 to give the idea. 

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(data.table)
library(DT)
combinations <- function(n) 
    expand.grid(rep(list(0:1),n))


pairwise <- combinations(6) 
pairwise$sum <- rowSums(pairwise)
pairwise <- subset(pairwise, pairwise$sum == 3)
colnames(pairwise) <- 1:ncol(pairwise)
rownames(pairwise) <- 1:nrow(pairwise)
pairwise <- pairwise[,1:6]

calcTestStat <- function(obs, poten) {
  testStat <- c()
    for (i in 1:nrow(poten)) {
      obs$W <- t(poten[i,])
      means <- obs %>% group_by(W) %>% 
      summarise_each(funs(sum(., na.rm=TRUE)), `Y(0)`, `Y(1)`)
      means <- means[, 2:3]
      means$average = rowSums(means) / (nrow(obs)/2)
      means <- as.matrix(means)
      diff <- abs(means[1,3] - means[2, 3])
      testStat <-c(testStat, diff)
    }  
  poten <- cbind(poten, testStat)
  return(poten)
}

tableWithTestStat <- calcTestStat(medical_example, pairwise)
datatable(head(tableWithTestStat), options = list(
  pageLength = 16, autoWidth = TRUE
))

```

But in a population of a larger size the the numbers quickly grow. 
For instance with 30 patients, we have ${30 \choose 15 } = 155,117,520$ number
of potential treatment plans. Of these plans the null hypothesis states that the 
observed outcome is as likely as any other outcome. 
More concretely we can derive a p-value (the probability of our observed 
case occuring given the null hypothesis) for our observed test-statistic from 
the fact that we can calculate all the potential test-statistics resulting from 
every possible treatment plan. Our p-value is the proportion of the possible treatment 
plans would result in test-statistic greater than or comparable to the observed test-statistic? 
A low p-value, typically below .05 indicates that the null hypothesis is false. 

This type of consideration allows us to assess whether there is an 
effect of the treatment, but not the degree to which the treatment is effective. 
Fortunately we can make use of the same set up to estimate the average causal 
effect to within an arbitary degree of confidence. 

Consider instead that we take a series of null hypotheses to be defined as: 

$Y_{c}(0) = Y_{t}(1) + c$ where $c \in C = [-3, -2.75, -2.5, ... 1.00]$

Then for each case we are able to fill in the outcomes table. 
If we let $c = .5$ then we have:

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(DT)

medical_example2 <- data.frame(
  unit <- c("Patient #1", "Patient #2", "Patient #3", "Patient #4", "Patient #5", "Patient  #6"),
  control_outcome <- c("(2.5)", "(4.5)", "(-0.5)", 4, 0, 1),
  treatment_outcome <- c(3, 5, 0, "(4.5)", "(.5)", "(1.5)"),
  treated <- c(1, 1, 1, 0, 0, 0),
  covariates <- c("3, 2, 1", "2, 4, 1", "5, 2, 4", "6, 1, 3", "4, 3, 2", "4, 1, 2")
)
colnames(medical_example2) <- c("unit", "Y(0)", "Y(1)", "W", "covariates")
datatable(medical_example2)
```

which gives us the following test-statistic. 

$T(\textbf{W}, \textbf{Y}^{obs}) = T^{diff} = 
|\overline{Y}^{obs}_{1} - \overline{Y}^{obs}_{0}| - c = |14.5/6 - 11.5/6| -.5 = 0$

we can now range over all the possible treatment plans and calculate a test statistic for each allowing us to derive a p-value for the observed result. Repeating this process for each $c \in C$ we gain a set of p-values which will circumscribe those elements of $C$ which have high positive probability under the assumed null hypothesis - giving us a range of values which are plausible candidates for the average causal effect of the treatment.
However to calculate the p-value under each possible treatment plan is computationally very expensive. Instead we approximate the pvalue for each hypothesis by sampling from the set of all possible random assignments. Consider a patient group of 14, then we have ${14 \choose 7} = 3432$ possible treatment plans. Our null hypothesis states that the treatment has a constant individual unit effect, so we can assess our observed t-statistic over the set of possible t-statistics derived from varying the treatment plan therby deriving a p-value for the observed result given the hypothesis of a constant effect. We can in the small cases where patients are a manageable number sample from the number of possible treatment plans to approximate the exact p-value. In the case of of 14 patients we can calculate the exact p-value, but with patient numbers 30+, we often need to use approximations.  If we iterate the process of series of constant values we can observe the candidate values which make our observed result more or less probable.   

```{r}
library(knitr)
library(DT)
library(xtable)
set.seed(1)

makeDF <- function(units) {
  len <- units
  Y0 <- rnorm(units, 0, 1)
  Y0[1:(len/2)] <- NA
  Y1 <- rnorm(units, 3, 1)
  Y1[((len/2)+1):len] <- NA
  units <- 1:len
  treatment <- c(rep(0, len))
  treatment[1: len/2] <- 1
  df <- data.frame(units, Y0, Y1, treatment)
  colnames(df) <- c("unit", "Y(0)", "Y(1)", "W")
  return(df)
}

potential_treatments <- c()
# Code for sampling when numbers are computationally too expensive.
# n <- NO. of units
#assignment <- rep(0:1, n/2)
#for(i in 1:10000) {
#  permutation <- sample(assignment)
#  if (!(list(permutation) %in% potential_treatments)) {
#  potential_treatments <- c(potential_treatments, list(permutation))
#  }
#}

combinations <- function(n) 
    expand.grid(rep(list(0:1),n))

comb <- combinations(14)
comb$sum <- rowSums(comb)
treatments <- subset(comb, comb$sum == 7)
treatments <- treatments[,1:14]
for (i in 1:nrow(treatments)) {
  potential_treatments <- c(potential_treatments, list(t(treatments[i,])))
}
print(length(potential_treatments))


calcTest <- function(df, candidate) {
  AvgT <- aggregate(`Y(1)` ~ W, df, mean)
  AvgT <- as.matrix(AvgT)
  AvgC <- aggregate(`Y(0)` ~ W, df, mean)
  AvgT <- as.matrix(AvgT)
  diff <- abs(AvgT[2, 2] - AvgC[1, 2] - candidate)
  return(diff)
}

calcPvalue <- function(df, potential_treatments, treatment, candidate) {
  df$W <- treatment
  df$`Y(1)` <- ifelse(is.na(df$`Y(1)`), df$`Y(0)` + candidate, df$`Y(1)`)
  df$`Y(0)` <- ifelse(is.na(df$`Y(0)`), df$`Y(1)` - candidate, df$`Y(0)`)
  observed <- calcTest(df, candidate)
  
  testStats <- c()
  for (j in 1:length(potential_treatments)) {
      df2 <- df
      df2$W <- potential_treatments[[j]]
      df2$`Y(1)` <- ifelse(is.na(df2$`Y(1)`), df2$`Y(0)` + candidate, df2$`Y(1)`)
      df2$`Y(0)` <- ifelse(is.na(df2$`Y(0)`), df2$`Y(1)` - candidate, df2$`Y(0)`)
      testStats <- c(testStats, calcTest(df2, candidate))
  }
  pvalue <- length(testStats[testStats >= observed]) / length(testStats)
  return(c(observed, pvalue))
}

calcAll <- function(units, potential_treatments, treatment, hypothetical_differences) {
  pvalues <- data.frame(Candidate <- c(), Observed <- c(), Pvalue <- c())
  df <- makeDF(units)
  for (i in 1:length(hypothetical_differences)) {
    pvalue  <- calcPvalue(df, potential_treatments, treatment, hypothetical_differences[i])
    pvalues <- rbind(pvalues, c(hypothetical_differences[i], pvalue[1], pvalue[2]))
  }
  return(pvalues)
}

hypothetical_differences <- seq(-4, 4, .5) 
treatment <- c(rep(0, 14))
treatment[1:7] <- 1
result <- calcAll(14, potential_treatments, treatment, hypothetical_differences)
colnames(result) <- c("Hypothetical Difference", "Observed TestStat", "P-value in Sample")
df <- makeDF(14)
kable(head(df, 14))
kable(result)
```